Vagueness, Noise, and Communication

### 2014-10-02 paper/01_intro.tex
introduction and background; sec com

\subsection{Sim-max games}

@@

Seen in this light, sim-max games appear to have the potential of
providing a foundation to approaches in cognitive semantics that rely
on the notion of conceptual spaces
\citep[e.g.][]{Gardenfors2000:Conceptual-Spac}.

\bigskip

\begin{itemize}
\item ESSs of sim-max games
\item Voronoi tessallations
\item conceptual spaces
\item desiderata on vagueness: higher order vagueness \dots
\item no vagueness normally, but possible Franke, O'Connor ...
\end{itemize}

\bigskip

@@

As in other applications of evolutionary game theory, we can
distinguish micro- and macro-level approaches. Micro-level approaches
look at adaptive behavior of individual agents. Usually, changes in
the behavioral dispositions of agents occur after every single
interaction. In contrast, macro-level approaches outline more
abstract, aggregate dynamics, happening in a huge population of
agents, or otherwise abstracting from seemingly irrelevant
detail. Usually, a macro-level dynamic captures changes of frequencies
of behavioral types in the population over time. Previous accounts
have suggested both micro- and macro-level dynamics to show how vague
signal use can arise in sim-max games.

As for micro-dynamics, \citet{FrankeJager2010:Vagueness-Signa} show
how limited memory of past interactions can lead to vague signal use,
when averaging over a single agent's behavior over time or over the
momentary behavior of a population of several language
users. \citet{OConnor2013:The-Evolution-o} introduces a variant of
reinforcement learning that includes a low-level form of stimulus
generalization. Agents update their behavior after each round of play
in such a way that similar states to the ones that actually occurred
are subject to behavioral adjustment as well. (Alternative approaches
are discussed in more detail in Section~XYZ, where we can compare them
with the system introduced in this paper.)

\citet{FrankeJager2010:Vagueness-Signa} also consider a macro-level
approach, using the notion of a \emph{quantal response}. A quantal
response function is a probabilistic choice rule that formalizes the
idea that agents make small mistakes when calculating the expected
utility of choice options. In aggregation, these probabilistic
mistakes lead to systematic ``trembles'' that produce vague signal
use. The approach we take here is superficially similar, but there are
crucial divergences. For one, we adopt a dynamic perspective by
looking at the stable states of a suitable population dynamics. For
another, we demonstrate in Section~XYZ that quantal responses can give
rise to counterintuitive predictions. These counterintuitive examples
suggest that vagueness in signaling use is not convincingly explained
by appeal to mistakes in calculating expected utility, but rather, as
we assume here, as the result of confusing similar states of the word.

Against this background, we introduce a new macro-level approach
to the evolution of vague signal use in sim-max games, where the
source of vagueness is the agents' natural disability to sharply
distinguish perceptually distinct states. This applies to both
production and comprehension, i.e., both to the sender as well as to
the receiver role.

We call our dynamic \textbf{replicator diffusion dynamic} (\rdd),
because it is a special case of the replicator mutator dynamic
(\rmd). It is a special case in the sense that the \rdd incorporates a
special kind of mutation that is based on the similarity of states (as
defined in the sim-max game). More concretely, we start from the idea
that similar states are more easily confused than less similar
states. To formalize this idea, we assume that there is a confusion
matrix $C$, where $C_{ij}$ is the probability that a state \mystate{i}
is realized as a possibly different state \mystate{j}. (There is a
slightly different sense of ``realization'' in the sender and the
receiver side, and we will enlarge on this below.) Based on this, we
show two things. For one, we show how the discrete-time formulation of
the replicator dynamic can be easily adapted to integrate diffusion of
behavior due to confusion of states. This yields the \rdd in its basic
discrete-time formulation. For another, we also show that there exists
a translation of the confusion matrix $C$ into a suitable mutation
matrix $M_C$ such that the effect of $C$ and $M_C$ are ``equivalent''
in a sense to be made precise. This supports the conclusion that the
\rdd can indeed be thought of as a special case of the \rmd.

>>>>

\subsection{Sim-max games \& conceptual spaces}

@@

Seen in this light, sim-max games appear to have the potential of
providing a foundation to approaches in cognitive semantics that rely
on the notion of conceptual spaces.
\citet[][70--77]{Gardenfors2000:Conceptual-Spac}, for example, has
prominently argued that natural categories are convex regions in a
suitable conceptual space. If the conceptual space has a suitable
metric, then it is possible to think of these convex categories as
derived from a set of prototypes. Fixing a set of prototypes, we
consider the category corresponding to each prototype $p$ as the set
of points that are more similar to $p$ than to any other
\citep[e.g.][]{OkabeBoots2000:Spatial-Tessell}. In this way,
\citet{Gardenfors2000:Conceptual-Spac} argues, an efficient
categorization system can be obtained: storing the prototypes lets us
recover the categories without having to store each categories
extension. However, what is left unexplained so far, is where the
prototypes come from, and why we would not see just any distribution
of prototypes as an equally efficient classification system. This is
where sim-max games, might chip in a principled approach to deriving
in an independent way, not only convex categories, but also
prototypical exemplars belonging to them.

@@

The brief outline of an approach to conceptual categorization using
sim-max games leaves many problems unaddressed. One of them is that
natural categories for continuously variable stimuli, like shades of
color, pitch heights, spatial dimension and the like, do not have
unique prototypes and clear category boundaries. We would like to
account for the possibility of vagueness, in particular the following
criteria
\citep[e.g.][]{Sainsbury1991:Is-There-Higher,KeefeSmith1997:Vagueness:-A-Re,Smith2008:Vagueness-and-D}:
(i) clear positive examples of a vague category should show a
gradient, perhaps smooth, transition to negative examples to
accommodate also ``higher-order vagueness'' ; (ii) prototypes should
likewise be gradient regions, peaking at the center of the vague
category they represent.

\citet{DouvenDecock2011:Vagueness:-A-Co} show that
\citeauthor{Gardenfors2000:Conceptual-Spac}'s conceptual spaces
approach can be extended to account for the existence of borderline
cases. From the assumption that prototypes are not unitary points, but
extended, yet convex regions in conceptual space,
\citeauthor{DouvenDecock2011:Vagueness:-A-Co} give a construction
algorithm that yields ``collated Voronoi diagrams'' with thick,
extended boundaries representing borderline regions. Building on this
work, \citet{DecockDouven2012:What-is-Graded-} show further how it is
possible to arrive at ``higher-order vagueness'' and degrees of
category membership, by, essentially, weighing in the distance of
different borderline examples to various prototypical regions. This
accounts for the first of the two desiderata mentioned above, but
still assumes that crisp non-gradient prototype regions must be
pre-given.

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games for certain types
of solution concepts, that can be divided into micro- and macro-level
approaches. Micro-level approaches look at adaptive behavior of
individual agents. Usually, changes in the behavioral dispositions of
agents occur after every single interaction. In contrast, macro-level
approaches outline more abstract, aggregate dynamics, happening in a
huge population of agents, or otherwise abstracting from seemingly
irrelevant detail. Usually, a macro-level dynamic captures changes of
frequencies of behavioral types in the population over time.

As for micro-dynamics, \citet{FrankeJager2010:Vagueness-Signa} show
how limited memory of past interactions can lead to vague signal use,
when averaging over a single agent's behavior over time or over the
momentary behavior of a population of several language
users. \citet{OConnor2013:The-Evolution-o} introduces a variant of
reinforcement learning that includes a low-level form of stimulus
generalization. Agents update their behavior after each round of play
in such a way that similar states to the ones that actually occurred
are subject to behavioral adjustment as well. (Alternative approaches
are discussed in more detail in Section~\ref{sec:discussion}, where we
can compare them with the system introduced in this paper.)

\citet{FrankeJager2010:Vagueness-Signa} also consider a macro-level
approach, using the notion of a \emph{quantal response}. A quantal
response function is a probabilistic choice rule that formalizes the
idea that agents make small mistakes when calculating the expected
utility of choice options. In aggregation, these probabilistic
mistakes lead to systematic ``trembles'' that produce vague signal
use. The approach we take here is superficially similar, but there are
crucial differences. For one, we adopt a dynamic perspective by
looking at limited-time outcomes of a dynamic process. For another, we
demonstrate in Section~\ref{sec:discussion} that quantal responses can
give rise to counterintuitive predictions. These counterintuitive
examples suggest that vagueness in signaling use is not convincingly
explained by appeal to mistakes in calculating expected utility, but
rather, as we assume here, as the result of confusing similar states
of the word.

Against this background, we introduce a new macro-level approach to
the evolution of vague signal use in sim-max games, where the source
of vagueness is the agents' natural inability to sharply distinguish
perceptually distinct states. This applies to the sender as well as to
the receiver role (albeit in slightly different manner). We call our
dynamic \emph{replicator diffusion dynamic} (\rdd), because it is a
special case of the replicator mutator dynamic (\rmd). It is a special
case in the sense that the \rdd incorporates a particular kind of
mutation that is based on the similarity of states (as defined in the
sim-max game). More concretely, we start from the idea that similar
states are more easily confused than less similar states. To formalize
this idea, we assume that there is a confusion matrix $C$, where
$C_{ij}$ is the probability that a state \mystate{i} is realized as a
possibly different state \mystate{j}. (There is a slightly different
sense of ``realization'' in the sender and the receiver side, and we
will enlarge on this below.) Based on this, we show two things. For
one, we show how the discrete-time formulation of the replicator
dynamic can be easily adapted to integrate diffusion of behavior due
to confusion of states. This yields the \rdd in its basic
discrete-time formulation. For another, we also show that there exists
a translation of the confusion matrix $C$ into a suitable mutation
matrix $M_C$ such that the effect of $C$ and $M_C$ are ``equivalent''
in a sense to be made precise. This supports the conclusion that the
\rdd can indeed be thought of as a special case of the \rmd.

Based on data from numerical simulations, we show that the inclusion
of mild levels of such stochastic noise in the replicator dynamics
does not disrupt the possibility of evolving communicative efficient
signaling strategies at all. On the contrary, there might even be a
higher-order benefit to the presence of perceptual imprecision, in
that it accelerates converges, but also unifies and regularizes
evolutionary outcomes in such a way that inefficient categorization is
avoided. These results complement similar results by
\citet{OConnor2013:The-Evolution-o}, obtained for a micro-dynamic
extension of reinforcement learning. The \rdd adds to this a more
general, abstract framework that is not tied to the specific
assumptions of turn-based reinforcement.
